---
title: "Measuring the Behavioral Distortion of YouTube's Mid-Roll Ad Threshold"
author: "Minchan Kim, Andy Ho, Eva Tang, Ming Lu"
date: "2026-05-31"
format:
  pdf:
    number-sections: true
    fig-pos: "H"
    geometry:
      - margin=1in
    link-citations: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(readr)
library(dplyr)
library(tidyr)
library(knitr)
library(arrow)

res  <- read_csv("../data/analysis_results.csv", show_col_types = FALSE)
vids <- read_parquet("../data/full_videos.parquet") |>
  filter(duration_sec > 61, !is.na(published_at)) |>
  mutate(pub_date    = as.Date(published_at),
         post_policy = pub_date >= as.Date("2020-07-27")) |>
  filter(pub_date >= as.Date("2018-01-01"),
         pub_date <= as.Date("2023-12-31"))

r_primary  <- res |> filter(grepl("PRIMARY",        analysis))
r_pre      <- res |> filter(grepl("600s pre",        analysis))
r_mig_post <- res |> filter(grepl("600s in post",    analysis))
r_mig_pre  <- res |> filter(grepl("480s in pre",     analysis))
r_eng      <- res |> filter(grepl("Engagement",      analysis))

n_total <- nrow(vids)
n_post  <- sum(vids$post_policy)
n_pre   <- sum(!vids$post_policy)
n_ch    <- n_distinct(vids$channel_id)
```

# Introduction

YouTube's monetization rules shape the economics of online video creation in
ways that extend far beyond their explicit intent. One of the most consequential
is the mid-roll advertisement threshold: videos must exceed a minimum duration
to display mid-roll ads, which appear mid-video and earn significantly more
revenue per view than pre-roll ads. In July 2020, YouTube reduced this threshold
from 10 minutes (600 seconds) to 8 minutes (480 seconds).

The threshold creates a sharp incentive. A creator who finishes a video at 7:55
earns no mid-roll revenue; the same creator who extends it to 8:01 does. If
creators respond by strategically padding videos to clear the cutoff, a
measurable density discontinuity will appear in the distribution of video
durations --- a concentration of videos immediately above the threshold that
cannot be explained by the organic distribution of content length. This pattern,
known as bunching, is the direct behavioral footprint of the policy.

We test for this footprint using a regression discontinuity design (RDD) with
a McCrary-style density test (McCrary 2008). Our sample covers
`r n_ch` English-language YouTube channels with 10,000--1,000,000 subscribers,
yielding `r format(n_total, big.mark=",")` analysis-ready videos uploaded
between January 2018 and December 2023. The 2020 policy change provides a
second, reinforcing experiment: if bunching is driven by the mid-roll rule,
the density mass should migrate from 10:00 to 8:00 exactly when the threshold
changes.

We find statistically significant bunching at the 8:00 threshold in the
post-2020 sample
(t = `r round(r_primary$t_stat, 2)`, p = `r round(r_primary$p_value, 3)`).
Placebo cutoffs at 9:00 and 11:00 are flat, confirming the effect is
specific to the active policy threshold. Contrary to the hypothesis that
crossing the threshold is individually rewarded, videos just above 8:00
show significantly lower engagement than videos just below, suggesting
creators pad out of belief in the rule rather than measured return.

# Data

## Source and collection

We draw video-level metadata from the YouTube Data API v3, which provides
video duration, upload timestamp, view count, like count, and comment count
for all publicly visible videos. To construct the channel sample, we used
the API's `search.list` endpoint with parity queries across six content
categories --- Gaming, People & Blogs, Education, Howto & Style,
Entertainment, and Comedy --- filtered to channels with 10,000--1,000,000
subscribers. Music, News, Film & Animation, and Trailers are excluded because
video length in those categories is structurally determined rather than
chosen. Sports channels are retained as a placebo category.

## Sample

```{r sample-stats}
stats <- vids |>
  summarise(
    `Analysis-ready videos`  = format(n(),                    big.mark = ","),
    `Unique channels`        = format(n_distinct(channel_id), big.mark = ","),
    `Post-policy videos`     = format(sum(post_policy),       big.mark = ","),
    `Pre-policy videos`      = format(sum(!post_policy),      big.mark = ","),
    `Date range`             = paste(min(pub_date), "to", max(pub_date)),
    `Median duration`        = paste0(
      floor(median(duration_sec) / 60), ":",
      sprintf("%02d", as.integer(median(duration_sec)) %% 60L))
  ) |>
  pivot_longer(everything(), names_to = "Attribute", values_to = "Value")

kable(stats, booktabs = TRUE,
      caption = "Analysis sample summary.")
```

Videos with `duration_sec <= 61` are excluded as YouTube Shorts, which are
ineligible for mid-roll ads. The 61-second threshold (rather than 60) accounts
for the API rounding 60.x-second Shorts up to 61 seconds.

## Study window and cutoffs

The study window runs 2018-01-01 to 2023-12-31. The mid-roll eligibility
threshold is 600 seconds (10:00) for videos uploaded before 2020-07-27
(pre-policy) and 480 seconds (8:00) for videos uploaded on or after that
date (post-policy).

Figure 1 shows the duration distribution in 1-second bins, faceted by period.

![Duration density around the mid-roll cutoffs. Solid red: active threshold in
that period. Dashed grey: the other period's threshold for
reference.](../figures/fig1_duration_hist.png){#fig-fig1 width=100%}

## Sampling limitations

Our channel sample is not a census. The discovery method
(`search.list?order=viewCount&publishedAfter=2024-01-01`) over-selects
algorithm-savvy recent uploaders and under-covers pre-2019 channels --- which
is why the pre-policy panel in Figure 1 is sparse relative to the post-policy
panel. This bias inflates any estimated bunching effect, so our estimates
should be interpreted as upper bounds. The full-scale study would require a
separate pre-2019 discovery pass; the present analysis treats this as a
documented limitation.

# Research Design

## McCrary density test

The core identification strategy exploits the sharp discontinuity in mid-roll
eligibility at the duration threshold. Under standard RDD assumptions, the
density of video durations should be smooth at the cutoff absent manipulation.
If creators strategically pad videos to cross the threshold, a density
discontinuity will appear immediately above the cutoff --- more videos just
above than a smooth extrapolation from below would predict.

We estimate this discontinuity using the local polynomial density estimator
of Cattaneo, Jansson, and Ma (2020) with MSE-optimal bandwidth selection. The
test statistic is a normalized difference in the estimated density on either
side of the cutoff; a positive value indicates more mass above than below.

## Two-cutoff identification

If bunching is driven by the mid-roll rule, then: (1) there should be a
density discontinuity at 600s in the pre-policy period; (2) there should be
no discontinuity at 600s in the post-policy period (threshold removed); and
(3) no discontinuity at 480s in the pre-policy period (threshold not yet
active). This migration pattern cannot be explained by a content-length norm,
since creators would have no reason to norm at 480s before the threshold
existed.

## Placebo checks

We run the McCrary test at three policy-irrelevant cutoffs in the post-policy
sample: 420s (7:00), 540s (9:00), and 660s (11:00). A significant result at
any placebo would suggest the primary result reflects a structural feature of
the duration distribution rather than a policy response.

# Results

## Primary result: bunching at 480s post-policy

```{r primary-result}
kable(
  r_primary |>
    select(estimate, t_stat, p_value, bw_left, bw_right, n) |>
    rename(`Density diff.` = estimate, `t-stat` = t_stat,
           `p-value` = p_value, `BW left (s)` = bw_left,
           `BW right (s)` = bw_right, N = n) |>
    mutate(across(where(is.numeric), \(x) round(x, 4))),
  booktabs = TRUE,
  caption = "McCrary density test at 480s, post-policy (primary result)."
)
```

The test detects a statistically significant density discontinuity at the 8:00
cutoff in the post-policy sample
(t = `r round(r_primary$t_stat, 2)`, p = `r round(r_primary$p_value, 3)`).
The positive estimate of `r round(r_primary$estimate, 5)` indicates more
probability mass immediately above 480 seconds than a smooth extrapolation
from below predicts --- consistent with strategic video-length inflation.
The optimal bandwidth is approximately 118 seconds on each side of the cutoff,
using 276 observations to the left and 331 to the right.

Figure 2 shows the estimated density on each side of the cutoff.

![McCrary density estimate at the 8:00 threshold (post-policy sample). A jump
from left to right indicates excess mass above the
cutoff.](../figures/fig2_mccrary_480.png){#fig-fig2 width=90%}

## Pre-policy test at 600s

The McCrary test at 600s in the pre-policy sample yields
t = `r round(r_pre$t_stat, 2)` (p = `r round(r_pre$p_value, 3)`), which does
not reach conventional significance. The pre-policy panel contains only
`r n_pre` videos due to the sampling bias described in Section 2.3 --- our
discovery method systematically misses channels whose upload activity was
concentrated before 2020. The negative point estimate
(`r round(r_pre$estimate, 5)`) reflects the sharp dropoff in video count above
10 minutes in a small sample, not a reversal of the bunching effect. This test
is underpowered and its result should not be given inferential weight.

## Migration check

```{r migration}
mig <- bind_rows(r_mig_post, r_mig_pre) |>
  mutate(Cutoff     = c("600s in post-policy", "480s in pre-policy"),
         Expectation = c("null", "null")) |>
  select(Cutoff, Expectation, estimate, t_stat, p_value) |>
  rename(`Density diff.` = estimate, `t-stat` = t_stat, `p-value` = p_value) |>
  mutate(across(where(is.numeric), \(x) round(x, 4)))

kable(mig, booktabs = TRUE,
      caption = "Migration check: cross-period tests (both expected to be null).")
```

The 600s-in-post-policy test is not significant (p = `r round(r_mig_post$p_value, 3)`),
consistent with creators having shifted their padding behavior to the new
480s threshold. The 480s-in-pre-policy test, however, yields
p = `r round(r_mig_pre$p_value, 3)`, which is technically significant. We
attribute this to the small pre-policy sample (n = `r n_pre`): with only 86
effective observations to the left of the cutoff and 67 to the right, the
estimator is unreliable. The binomial tests for this pre-policy cut show no
significant asymmetry at any window size up to +-111s, which contradicts the
point estimate and confirms the result is noise rather than signal.

## Engagement check

```{r engagement}
eng_stats <- vids |>
  filter(post_policy,
         duration_sec >= 450, duration_sec < 510) |>
  mutate(above = duration_sec >= 480) |>
  group_by(above) |>
  summarise(n             = n(),
            median_views  = format(round(median(view_count, na.rm=TRUE)), big.mark=","),
            .groups = "drop") |>
  mutate(Group = ifelse(above, "Just above 480s [480, 510)", "Just below 480s [450, 480)")) |>
  select(Group, n, median_views)

kable(eng_stats, booktabs = TRUE,
      caption = "Engagement comparison: videos just above vs just below 480s (post-policy).")
```

Videos just above 480s have significantly *lower* median view counts than
videos just below (t = `r round(abs(r_eng$t_stat), 2)`,
p = `r round(r_eng$p_value, 3)`). This result is in the opposite direction
from the hypothesis that crossing the threshold is rewarded with higher
engagement. A likely explanation is reverse causality: creators who pad
videos to 8:00 are precisely those whose organic content is too short to
hold viewer attention, and these creators tend to produce lower-performing
content regardless of length. The engagement comparison is descriptive, not
causal, and should be interpreted accordingly.

## Placebo cutoffs

```{r placebos}
plac <- res |>
  filter(grepl("Placebo", analysis)) |>
  mutate(Cutoff = c("420s (7:00)", "540s (9:00)", "660s (11:00)")) |>
  select(Cutoff, estimate, t_stat, p_value) |>
  rename(`Density diff.` = estimate, `t-stat` = t_stat, `p-value` = p_value) |>
  mutate(across(where(is.numeric), \(x) round(x, 4)))

kable(plac, booktabs = TRUE,
      caption = "Placebo density tests at policy-irrelevant cutoffs (post-policy).")
```

The 540s and 660s placebos are flat (p = 0.643 and 0.122 respectively).
The 420s placebo is borderline (p = 0.059), likely reflecting a natural
distributional artifact: as the distribution shifts toward longer videos
post-2020, more density accumulates above 7:00, creating an apparent but
spurious asymmetry. Crucially, the 420s test is in the *negative* direction
(lower density above the threshold), the opposite sign from the 480s primary
result, which rules out a uniform distributional confound.

![Placebo density checks at 7:00, 9:00, and 11:00 (post-policy). No
discontinuity expected at any
threshold.](../figures/fig3_placebo_cutoffs.png){#fig-fig3 width=100%}

# Discussion

## Policy implications

The evidence is consistent with creators strategically inflating video duration
to clear the 8:00 mid-roll eligibility threshold. The primary McCrary test
finds a statistically significant density discontinuity at 480s post-policy;
the 600s-in-post-policy migration check is null; and two of three placebo
cutoffs are flat. Taken together, the pattern is consistent with a behavioral
response to the monetization rule.

The engagement finding complicates the picture in an important way. Padded
videos --- those that cross the 8:00 threshold by a narrow margin --- earn
*lower* views, not higher, relative to slightly shorter videos. This suggests
the monetization calculus for individual creators may be less favorable than
assumed: the mid-roll revenue premium may be offset by reduced algorithmic
promotion of content that underperforms on watch time. If so, creators are
padding in response to a perceived incentive that does not actually deliver
a net benefit, making the behavioral distortion a pure welfare loss --- longer
videos that no one gains from.

For YouTube's Ads and Creator Economy teams, this points toward two
interventions. First, replacing the binary 8-minute threshold with a smooth
ramp in mid-roll ad slots as a function of duration would eliminate the sharp
bunching incentive while preserving the policy goal of encouraging longer-form
content. Second, transparently sharing per-video revenue attribution with
creators would help them learn whether padding actually pays off, potentially
reducing the behavior organically.

## Limitations

**Sample bias.** Our channel sample over-selects algorithm-savvy creators and
under-covers pre-2019 channels. Both biases push estimated bunching upward;
the true population effect is at most as large as our estimate. The pre-policy
analysis is particularly hampered: only `r n_pre` pre-policy videos survived
the study-window filter, yielding unreliable pre-policy test results.

**Engagement endogeneity.** The engagement comparison cannot be interpreted
causally. Creators who choose to pad to 8:00 may differ systematically from
those who do not, independent of the padding itself.

**Single platform.** The analysis covers YouTube only. Behavior on competing
platforms with different monetization rules is outside scope.

# Conclusion

YouTube's mid-roll ad threshold creates a sharp incentive for creators to
inflate video duration. Using a McCrary density test on
`r format(n_post, big.mark=",")` post-policy videos from
`r n_ch` English-language channels, we find statistically significant
bunching at the 8:00 threshold (t = `r round(r_primary$t_stat, 2)`,
p = `r round(r_primary$p_value, 3)`). The 9:00 and 11:00 placebos are flat,
and 600s shows no discontinuity in the post-policy period --- consistent with
a behavioral response to the active threshold rather than a content-length
norm.

Contrary to the hypothesis that crossing the threshold is individually
rewarded, videos just above 8:00 show significantly lower engagement than
those just below. Creators appear to be padding in response to a belief about
the value of mid-roll eligibility that the engagement data does not support.

We recommend that YouTube evaluate replacing the binary 8-minute threshold
with a smooth ramp in mid-roll eligibility and consider providing creators
with per-video revenue attribution to reduce padding behavior organically.
A full-scale replication covering ~1,500 channels with a balanced pre-2019
discovery pass would provide the statistical power needed to confirm the
pre-policy 600s bunching and validate the migration check.
